---
title: "Chapter 10 Causality and Matching"
format: 
    html:
        number-sections: true
---

## Philosophical Aspects on Causality

## Causal Inference

1. graphical models
2. predictive causality
3. counterfactual causality

Dynamic Path Analysis

### Graphic Models

Wright (1921)

### Predictive Causality

stochastic processes

__Granger Test__

__Local Dependence__

### Counterfactuals

- Rubin school: matching
- Robins school: inverse probability weighting

## Aalen's Additive Hazards Model

for proportional hazards model

$$h(t|x) = h_0(t)r(\beta, x_i(t)), \quad t \gt 0 $$

additive hazards model

$$h(t|x_i) = h_0(t)+\beta_1(t)x_{i1}(t)+\ldots+\beta_p(t)x_{ip}(t), \quad t \gt 0$$

```{r}
library(eha)
library(survival)
```

```{r aareg}
fit = aareg(formula = Surv(enter-60, exit-60, event) ~ sex, data = oldmort)
summary(fit)
```

```{r plot}
oldpar <- par(mfrow = c(1, 2))
plot(fit)
par(oldpar)
```

## Dynamic Path Analysis

## Matching

### Paired Data

### More than one control

## Conclusion

- Causal thinking is important
- Counterfactual reasoning and marginal models yield little insight into "how it works"
- Creation of pseudo-populations through weighting and matching may limit the understanding of how things really work
    - Analyze the process as it presents itself, so that it is easier to generalize findings

