---
title: "chapter 6 more on cox regression"
format:
    html:
        number-sections: true
---

## Stratification

??test for proportionality assumption

```{r}
library(eha)

fert1 <- fert[fert$parity == 1, ]
levels(fert1$parish) <- c("Jorn", "Norsjo", "Skelleftea")
fert1$parish <- relevel(fert1$parish, ref = "Skelleftea")
fit <- coxreg(Surv(next.ivl, event) ~ strata(ses) + I(age - 25) + I(year-1860) + prev.ivl + parish, data = fert1)
```

```{r}
par(lwd=1.5, cex=0.8)
plot(fit, ylab = "Cumulative hazards", xlab = "Years")
```

## Sampling of Risk Sets
```{r}
fit <- coxreg(Surv(enter, exit, event) ~ ses, data = mort)

fit.4 <- coxreg(Surv(enter, exit, event) ~ ses, data = mort, max.survs = 4)

f1 <- coefficients(summary(fit))[c(1, 3)]
f4 <- coefficients(summary(fit.4))[c(1, 3)]

out <- rbind(f1, f4)
colnames(out) <- c("Coef", "se(Coef)")
rownames(out) <- c("Original", "Sample")
round(out, 4)
```


## Residuals

```{r}
library(survival)
head(kidney)

kl <- kidney[!duplicated(kidney$id), ]

fit <- coxreg(Surv(time, status) ~ disease + age + sex, data = kl)

plot(residuals(fit))
```

## checking model assumptions

### Proportionality

```{r}
library(eha)
fert2 <- fert[fert$parity == 2, ]
fit <- survival::coxph(Surv(next.ivl, event) ~ ses + age + year + parish, data = fert2)

prop.full <- survival::cox.zph(fit)
prop.full
```


```{r}
fit1 <- coxph(Surv(next.ivl, event) ~ strata(ses) + age + year + parish, data = fert2)
prop.full <- survival::cox.zph(fit1)
prop.full
```

So the `age` variable does not conform to the proportionality constraint. Let check its distribution and try to categorise it.

```{r}
hist(fert2$age, main="", xlab="age")
```

```{r}
fert2$qage <- cut(fert2$age, 4)
fit2 <- coxph(Surv(next.ivl, event) ~ strata(ses) + qage + year + parish, data = fert2)

fit2.zph <- survival::cox.zph(fit2)
fit2.zph
```

### log-linearity
```{r}
plot(fert2$age, residuals(fit))
lines(lowess(fert2$age, residuals(fit)))
```

## Fixed Study Period Survival
Basically what we do in financial risk modelling is a fixed study period survival analysis. The behavioural period is the fixed observational period.

## The Weird Bootstrap

This bootstrap took too long time to run. So I commented it out.


```{r}
# fit <- coxreg(Surv(enter, exit, event) ~ sex, boot=300, data = child)
# b_hat <- fit$coefficients[1]
# b_se <- sqrt(fit$var[1, 1])
# b_sample <- fit$bootstrap[1, ]
```

```{r}
# plot(density(b_sample - b_hat, bw = 0.005), xlab = "bootstrap - beta_hat", main="")
# abline(h=0, v=0)
```
